\title{Data Types, Functions, and Programming}
\titlerunning{Data Types, Functions, and Programming}
\toctitle{Data Types, Functions, and Programming}
\author{Daniel R. Grayson%
        \thanks{Supported by NSF grant DMS 99-70085.}
        % \inst 1
   \and Michael E. Stillman%
        %\inst 2
        %\fnmsep
        \thanks{Supported by NSF grant 99-70348.}}
\authorrunning{D. R. Grayson and M. E. Stillman}
% \institute{University of Illinois at Urbana-Champaign, Department of
%   Mathematics, Urbana, IL 61801, USA
%       \and Cornell University, Department of Mathematics, Ithaca, NY 14853, USA}

\maketitle

\begin{abstract}
  In this chapter we present an introduction to the structure of \Mtwo
  commands and the writing of functions in the \Mtwo language.  For further details
  see the \Mtwo manual distributed with the program~\cite{M2}.
\end{abstract}

\section{Basic Data Types}

The basic \ie{data types} of \Mtwo include numbers of various types (integers,
rational numbers, floating point numbers, complex numbers), lists (basic
lists, and three types of visible lists, depending on the delimiter used),
hash tables, strings of characters (both 1-dimensional and 2-dimensional),
Boolean values (true and false), symbols, and functions.  Higher level types
useful in mathematics are derived from these basic types using facilities
provided in the \Mtwo language.  Except for the simplest types (integers and
Boolean values), \Mtwo normally displays the type of the output value on a
second labeled output line.

Symbols\index{symbols} have a name which consists of letters, digits, or apostrophes, the
first of which is a letter.  Values can be assigned to symbols and recalled
later.
<<<w>>>
<<<w = 2^100>>>
<<<w>>>
Multiple values can be assigned in parallel.
<<<(w,w') = (33,44)>>>
<<<w>>>
<<<w'>>>
Comments are initiated by {\tt --} and extend to the end of the line.
<<<(w,w') = (33,   -- this is a comment
          44)>>>
Strings\index{strings} of characters are delimited by quotation marks.
<<<w = "abcdefghij">>>
They may be joined horizontally to make longer strings, or vertically to make
a two-dimensional version called a {\sl net}.
<<<w | w>>>
<<<w || w>>>
Nets are used in the preparation of two dimensional output for polynomials.

Floating point numbers are distinguished from integers by the presence of a
decimal point, and rational numbers are entered as fractions.
<<<2^100>>>
<<<2.^100>>>
<<<(36 + 1/8)^6>>>
Parentheses, braces, and brackets are used as delimiters for the three types
of {\em \ie{visible lists}}: \ie{lists}, \ie{sequences}, and \ie{arrays}.
<<<x1 = {1,a}>>>
<<<x2 = (2,b)>>>
<<<x3 = [3,c,d,e]>>>
Even though they use braces, lists should not be confused with sets, which will be treated later.
A double period can be used to construct a sequence of consecutive elements in various contexts.
<<<1 .. 6>>>
<<<a .. f>>>
Lists can be nested.
<<<xx = {x1,x2,x3}>>>
The number of entries in a list is provided by {\tt \#}.
<<<#xx>>>
The entries in a list are numbered starting with 0, and can be recovered with
{\tt \#} used as a binary operator.
<<<xx#0>>>
<<<xx#0#1>>>
We can join visible lists and {\sl append} or {\sl prepend} an element to a visible
list.  The output will be the same type of visible list that was provided in
the input: a list, a sequence, or an array; if the arguments are various
types of lists, the output will be same type as the first argument. 
<<<join(x1,x2,x3)>>>\index{lists!joining}\indexcmd{join}%
<<<append(x3,f)>>>\index{lists!appending}\indexcmd{append}%
<<<prepend(f,x3)>>>\index{lists!prepending}\indexcmd{prepend}%
Use {\tt sum} or {\tt product} to produce the sum or product of all the
elements in a list.
<<<sum {1,2,3,4}>>>\indexcmd{sum}%
<<<product {1,2,3,4}>>>\indexcmd{product}%


\section{Control Structures}

Commands for later execution are encapsulated in {\sl \ie{functions}}.  A function
is created using the operator {\tt -\char`\>}\indexcmd{->} to separate the parameter or
sequence of parameters from the code to be executed later.  Let's try an
elementary example of a function with two arguments.
<<<f = (x,y) -> 1000 * x + y>>>
The parameters {\tt x} and {\tt y} are symbols that will acquire a
value later when the function is executed.  They are {\sl local}\index{variables!local} in the sense that
they are completely different from any symbols with the same name that occur
elsewhere.  Additional local variables for use within the body of a function
can be created by assigning a value to them with {\tt \char`\:\char`\=}
\indexcmd{:=}
(first time only).  We illustrate this by rewriting the function above.
<<<f = (x,y) -> (z := 1000 * x; z + y)>>>
Let's apply the function to some arguments.
<<<f(3,7)>>>
The sequence of arguments can be assembled first, and then passed to the
function.
<<<s = (3,7)>>>
<<<f s>>>
As above, functions receiving one argument may be called without parentheses.
<<<sin 2.1>>>
A compact notation for functions makes it convenient to apply them
without naming them first.  For example, we may use \indexcmd{apply}{\tt apply} to apply a
function to every element of a list and to collect the results into a list.
<<<apply(1 .. 10, i -> i^3)>>>
The function \indexcmd{scan}{\tt scan} will do the same thing, but discard the results.
<<<scan(1 .. 5, print)>>>
Use \indexcmd{if}{\tt if ... then ... else ...} to perform alternative actions
based on the truth of a condition.
<<<apply(1 .. 10, i -> if even i then 1000*i else i)>>>
A function can be terminated prematurely with \indexcmd{return}{\tt return}.
<<<apply(1 .. 10, i -> (if even i then return 1000*i; -i))>>>
Loops in a program can be implemented with \indexcmd{while}{\tt while ... do ...}.
<<<i = 1; while i < 50 do (print i; i = 2*i)>>>
Another way to implement loops is with \indexcmd{for}{\tt for} and
\indexcmd{do}{\tt do} or {\tt list}, with optional
clauses introduced by the keywords {\tt from}, {\tt to}, and {\tt when}.
<<<for i from 1 to 10 list i^3>>>
<<<for i from 1 to 4 do print i>>>
A loop can be terminated prematurely with \indexcmd{break}{\tt break}, which accepts an
optional value to return as the value of the loop expression.
<<<for i from 2 to 100 do if not isPrime i then break i>>>
If no value needs to be returned, the condition for continuing can be
provided with the keyword {\tt when}; iteration continues only as long as the
predicate following the keyword returns {\tt true}.
<<<for i from 2 to 100 when isPrime i do print i>>>


% if then else, while do, for, break, return,

\section{Input and Output}

The function \indexcmd{print}{\tt print} can be used to display something on the screen.
<<<print 2^100>>>
For example, it could be used to display the elements of a list on separate
lines.
<<<(1 .. 5) / print;>>>
The operator {\tt <<} can be used to display something on the
screen, without the newline character.
<<<<< 2^100>>>
Notice the value returned is a {\em file}\index{files}.  A {\em file} in \Mtwo is a data type that
represents a channel through which data can be passed, as input, as
output, or in both directions.  The file \indexcmd{stdio}{\tt stdio} encountered above
corresponds to your shell window or terminal, and is used for two-way
communication between the program and the user.  A file may correspond
to what one usually calls a file, i.e., a sequence of data bytes associated
with a given name and stored on
your disk drive.  A file may also correspond to a {\em socket}, a
channel for communication with other programs over the network.

Files can be used with the binary form
of the operator {\tt \char`\<\char`\<} to display something else on the same
line.
<<<<< "the value is : " << 2^100>>>
Using \indexcmd{endl}{\tt endl} to represent the new line character or character sequence, we can
produce multiple lines of output.
<<<<< "A = " << 2^100 << endl << "B = " << 2^200 << endl;>>>
We can send the same output to a disk file named {\tt foo}, but we must remember to
close it with {\tt close}.
<<<"foo" << "A = " << 2^100 << endl << close>>>
The contents of the file can be recovered as a string with \indexcmd{get}{\tt get}.
<<<get "foo">>>
If the file contains valid \Mtwo commands, as it does in this case, we can
execute those commands with \indexcmd{load}{\tt load}.
<<<load "foo">>>
We can verify that the command took effect by evaluating {\tt A}.
<<<A>>>
Alternatively, if we want to see those commands and the output they produce,
we may use \indexcmd{input}{\tt input}.
<<<input "foo">>>
Let's set up a ring for computation in \Mtwo.
<<<R = QQ[x,y,z]>>>\index{ring!making one}%
<<<f = (x+y)^3>>>
Printing, and printing to files, works for polynomials,
too.\index{printing!to a file}%
<<<"foo" << f << close;>>>
The two-dimensional output is readable by humans, but is not easy to convert
back into a polynomial.
<<<get "foo">>>
Use \indexcmd{toString}{\tt toString} to create a 1-dimensional form of the polynomial
that can be stored in a file in a format readable by \Mtwo and by other
symbolic algebra programs, such as {\em Mathematica} or {\em Maple}.
<<<toString f>>>
Send it to the file.
<<<"foo" << toString f << close;>>>
Get it back.
<<<get "foo">>>
Convert the string back to a polynomial with \indexcmd{value}{\tt value}, using \indexcmd{oo}{\tt oo} to
recover the value of the expression on the previous line.
<<<value oo>>>
The same thing works for matrices, and a little more detail is provided by
\indexcmd{toExternalString}{\tt toExternalString}, if needed.
<<<vars R>>>
<<<toString vars R>>>
<<<toExternalString vars R>>>

\section{Hash Tables}

Recall how one sets up a quotient ring for computation in \Mtwo.
<<<R = QQ[x,y,z]/(x^3-y)>>>
<<<(x+y)^4>>>
How does \Mtwo represent a ring like $R$ in the computer?  To answer that,
first think about what sort of information needs to be retained about $R$.
We may need to remember the coefficient ring of $R$, the names of the
variables in $R$, the monoid of monomials in the variables, the degrees of
the variables, the characteristic of the ring, whether the ring is
commutative, the ideal modulo which we are working, and so on.  We also may
need to remember various bits of code: the code for performing the basic
arithmetic operations, such as addition and multiplication, on elements of
$R$; the code for preparing a readable representation of an element of $R$,
either 2-dimensional (with superscripts above the line and subscripts below),
or 1-dimensional.  Finally, we may want to remember certain things that take
a lot of time to compute, such as the Gr\"obner basis of the ideal.

A {\sl \ie{hash table}} is, by definition, a way of representing (in the computer)
a function whose domain is a finite set.  In \Mtwo, hash tables are extremely
flexible: the elements of the domain (or {\sl keys\index{keys of a hash table}}) and the elements of the
range (or {\sl values\index{values of a hash table}}) of the function may be any of the other objects
represented in the computer.  It's easy to come up with uses for functions
whose domain is finite: for example, a monomial can be represented by the
function that associates to a variable its nonzero exponent; a polynomial can
be represented by a function that associates to a monomial its nonzero
coefficient; a set can be represented by any function with that set as its
domain; a (sparse) matrix can be represented as a function from pairs of
natural numbers to the corresponding nonzero entry.

Let's create a hash table and name it.
<<<f = new HashTable from { a=>444, Daniel=>555, {c,d}=>{1,2,3,4}}>>>
The operator \indexcmd{=>}{\tt \char`\=\char`\>} is used to represent a key-value pair.
We can use the operator \indexcmd{\#}{\tt \#} to recover the value from the key.
<<<f#Daniel>>>
<<<f#{c,d}>>>
If the key is a symbol, we can use the operator \indexcmd{.}{\tt .} instead; this is
convenient if the symbol has a value that we want to ignore.
<<<Daniel = a>>>
<<<f.Daniel>>>
We can use \indexcmd{\#?}{\tt \#?} to test whether a given key occurs in the hash table.
<<<f#?a>>>
<<<f#?c>>>
Finite sets are implemented in \Mtwo as hash tables: the elements of the set
are stored as the keys in the hash table, with the accompanying values all
being $1$.  (Multisets are implemented by using values larger than $1$, and
are called {\sl tallies}.)
<<<x = set{1,a,{4,5},a}>>>\indexcmd{set}%
<<<x#?a>>>
<<<peek x>>>
<<<y = tally{1,a,{4,5},a}>>>
<<<y#a>>>
We might use \indexcmd{tally}{\tt tally} to tally how often a function attains its various
possible values.  For example, how often does an integer have 3 prime
factors?  Or 4?  Use \indexcmd{factor}{\tt factor} to factor an integer.
<<<factor 60>>>
Then use {\tt \#} to get the number of factors.
<<<# factor 60>>>
Use {\tt apply} to list some values of the function.
<<<apply(2 .. 1000, i -> # factor i)>>>
Finally, use {\tt tally} to summarize the results.
<<<tally oo>>>

Hash tables turn out to be convenient entities for storing odd bits and
pieces of information about something in a way that's easy to think about and
use.
In \Mtwo, rings are represented as hash tables, as are ideals, matrices,
modules, chain complexes, and so on.  For example, although it isn't a
documented feature, the key {\tt ideal} is used to preserve the ideal that
was used above to define the quotient ring {\tt R}, as part of the
information stored in {\tt R}.  
<<<R.ideal>>>
The preferred and documented way for a user to recover this information is
with the function \indexcmd{ideal}{\tt ideal}. 
<<<ideal R>>>
Users who want to introduce a new high-level mathematical
concept to \Mtwo may learn about hash tables by referring to the \Mtwo manual
\cite{M2}.

\section{Methods}

You may use the \texttt{code}\indexcmd{code} command to locate the source code for a given
function, at least if it is one of those functions written in the \Mtwo
language.  For example, here is the code for {\tt demark}, which may be used
to put commas between strings in a list.
<<<code demark>>>
The code for tensoring a ring map with a module can be displayed in this way.
<<<code(symbol **, RingMap, Module)>>>\indexcmd{symbol}
The code implementing the {\tt ideal} function when applied to a
quotient ring can be displayed as follows.
<<<code(ideal, QuotientRing)>>>
Notice that it uses the key {\tt ideal} to extract the information from the
ring's hash table, as you might have guessed from the previous discussion.
The bit of code displayed above may be called a {\sl method}\index{method} as
a way of indicating that several methods for dealing with various types of
arguments are attached to the function named {\tt ideal}.  New such {\sl
  method functions} may be created with the function {\tt method}\indexcmd{method}.  Let's
illustrate that with an example: we'll write a function called {\tt denom}
which should produce the denominator of a rational number.  When applied to
an integer, it should return 1.  First we create the method function.
<<<denom = method();>>>
Then we tell it what to do with an argument from the class {\tt QQ} of rational numbers.
<<<denom QQ := x -> denominator x;>>>
And also what to do with an argument from the class {\tt ZZ} of integers.
<<<denom ZZ := x -> 1;>>>
Let's test it.
<<<denom(5/3)>>>
<<<denom 5>>>

\section{Pointers to the Source Code}

A substantial part of \Mtwo is written in the same language provided to the
users.  A good way to learn more about the \Mtwo language is to peruse the
source code that comes with the system in the directory {\tt Macaulay2/m2}.
Use the {\tt code} function, as described in the previous section, for
locating the bit of code you wish to view.

The source code for the interpreter of the \Mtwo language is in the directory
{\tt Macaulay2/d}.  It is written in another language designed to be mostly
type-safe, which is translated into {\tt C} by the translator whose own
{\tt C} source code is in the directory {\tt Macaulay2/c}.  Here is a
sample line of code from the file {\tt Macaulay2/d/tokens.d}, which shows how
the translator provides for allocation and initialization of dynamic data
structures.
\par
\vskip 5 pt
\begingroup
\tteight
\baselineskip=8pt
\lineskip=0pt
\obeyspaces
globalFrame := Frame(dummyFrame,globalScope.seqno,Sequence(nullE));\leavevmode\hss\endgraf
\endgroup
\penalty-1000
\par
\vskip 1 pt
\noindent
And here is the {\tt C} code produced by the translator.
\par
\vskip 5 pt
\begingroup
\tteight
\baselineskip=\outputBaseLineSkip
\lineskip=0pt
\obeyspaces
\obeylines
tokens\char`\_Frame tokens\char`\_globalFrame;
tokens\char`\_Frame tmp\char`\_\char`\_23;
Sequence tmp\char`\_\char`\_24;
tmp\char`\_\char`\_24 = (Sequence) GC\char`\_MALLOC(sizeof(struct S259\char`\_)+(1-1)*sizeof(Expr));
if (0 == tmp\char`\_\char`\_24) outofmem();
tmp\char`\_\char`\_24->len\char`\_ = 1;
tmp\char`\_\char`\_24->array\char`\_[0] = tokens\char`\_nullE;
tmp\char`\_\char`\_23 = (tokens\char`\_Frame) GC\char`\_MALLOC(sizeof(struct S260\char`\_));
if (0 == tmp\char`\_\char`\_23) outofmem();
tmp\char`\_\char`\_23->next = tokens\char`\_dummyFrame;
tmp\char`\_\char`\_23->scopenum = tokens\char`\_globalScope->seqno;
tmp\char`\_\char`\_23->values = tmp\char`\_\char`\_24;
tokens\char`\_globalFrame = tmp\char`\_\char`\_23;
\endgroup
\penalty-1000
\par
\vskip 1 pt

The core algebraic algorithms constitute the {\sl engine} of \Mtwo and are
written in {\tt C\char`\+\char`\+}, with the source files in the directory
{\tt Macaulay2/e}.  In the current version of the program, the interface
between the interpreter and the core algorithms consists of a single
two-directional stream of bytes.  The manual that comes with the system
\cite{M2} describes the engine communication protocol used in that
interface.

%% Mike says we don't need this:

%% One feature of \Mtwo which should be understood by system administrators, at
%% least of Unix systems, is the {\tt dumpdata} routine.  Since so much of the
%% source code is written in the interpreted \Mtwo language, and the task of
%% reading and parsing it can take an appreciable amount of time on slower
%% machines, we perform that task in advance (by the {\tt setup} command) when
%% the program is installed on your system.  The results are saved by writing
%% all of the data areas of the memory of the program to disk.  When a user runs
%% {\tt M2}, the file containing the saved contents is quickly mapped into
%% memory using the {\tt loaddata} command.  The contents of memory depend
%% in a detailed way on the architecture of the computer and the shareable
%% libraries that are installed, so when those things change, the {\tt setup}
%% command may need to be run again.  If those things change too often, the
%% system administrator may disable that feature.


%% Date: Tue, 21 Nov 2000 11:36:16 +0900 (JST)
%% From: Nobuki Takayama <taka@math.kobe-u.ac.jp>
%% To: dan@math.uiuc.edu
%% CC: mike@polygon.math.cornell.edu, taka@math.kobe-u.ac.jp
%% In-reply-to: <200011161511.JAA18405@orion.math.uiuc.edu> (dan@math.uiuc.edu)
%% Subject: Re: software
%% 
%% Dear Dan;
%% I would like to know 
%%   (1) "hints" to read and understand the M2 source code
%% and
%%   (2) what are new ideas in the M2 implementations.
%% The source code contains all ideas in detail, but it is not easy to
%% read and understand it without some hints.
%% 
%% I look forward to reading the section and the source code with the help
%% of the section.
%% Nobuki
%% 
%% >    If you want to add one subsection at the end, it could be something
%% >    like "Developers' notes" or "System Issues". The reader I have in mind
%% >
%% >    chapter is just the place. If you wish to consider this, then I suggest
%% >    that Mike communicate directly with Nobuki about the contents.
%% >
%% >Could you tell us directly whay you have in mind about that?
%% 
